\section{Implementation and Evaluation}
\label{sec:implementation}

We have implemented the prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml. Our implementation is called \namec. The input to \namec is a
program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
includes assignments, conditionals, loops, more primitive datatypes
(\eg, integers), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference and constraint solving
closely follows the description given in
Sec.~\ref{sec:type-inference}. The one difference is that our
constraint solver uses an approximation algorithm to compute the
minimal graph augmentation (\S~\ref{sec:csolve}).
% To solve the
% constraints that arise during type inference, we built a solver called
% \csolve that implements constraint solving approach based on fixpoint
% computation and graph augmentation described in \S~\ref{sec:csolve}. 
% Since graph augmentation is NP-hard, \csolve uses an approximation
% algorithm that repeatedly adds an edge and recomputes paths in graph
% $G_1$ until it is as connected as graph $G_2$. Though this approach
% may not necessarily return the weakest solution to the abduction
% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
% constraints are not often complex. 
% If the input $\absof{\FB^+}$ program does not create any unsafe
% references, \namec annotates it with region types, which act as a
% witness to program's memory safety. If potential safety violations are
% encountered, then the type inference fails during the constraint
% solving phase. 

\begin{figure}
\begin{codejava}
class LinkedList<T><R5,R4 | R4$\outlives$R5> {
  ListNode<T><R5,R4> head;
  ..
  List<T><R17,R4> rev<R17,R4 | R4$\outlives$R17>(unit u) {
    List<T><R17,R4> xs = 
        new List<T><R17,R4>(this.head.val);
    ListNode<T><R5,R4> cur = this.head.next;
    while (!cur == Null) {
      xs.add<R17>(cur.val)
      cur = cur.next;
    }
    return xs;
  }
\end{codejava}

\caption{Region-annotated definition of \C{rev} computed by \namec}
\label{fig:rev}
\vspace*{-0.15in}
\end{figure}

To evaluate the practical utility of our region type system and type
inference, we 
% implemented some of the C\# libraries in $\absof{\FB^+}$, and 
performed two kinds of experiments. First, we implemented some of the
microbenchmarks ($\le$100 LOC), which are the standard libraries such
as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
our inference engine to infer their principal region types. Since a
library class is region-oblivious, if it is well-typed as per the core
type system, then \namec must be able to automatically construct its
region-type-annotated definition without fail. As expected, \namec was
able to infer principal region types for all the library classes,
under 10ms. Fig~\ref{fig:rev} shows the region-type-annotated
definition computed for the list reverse method. Observe that \namec
was able to infer that the list and its data (of type \C{T}) can be
allocated in different regions, as long as the later outlives the
former. This allows, for instance, a \C{preOrder} method to traverse a
tree in a transferable region, and return a list of its nodes, where
the list itself is allocated in the stack region. 

Next, we translated some of the Naiad streaming query operator
benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
$\absof{\FB^+}$, and used \namec to verify their safety. During the
process, we found multiple instances of potential memory safety
violations in the $\absof{FB^+}$ translation of benchmarks, which we
verified to be present in the original C\# implementation as well. The
cause of all safety violations is the creation of a reference from the
outgoing message (a transferable region) to the payload of the
incoming message. For example, the implementation of \C{SelectVertex}
contains the following:
\begin{codejava}
  if (this.selector(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references. 

\dv{Hmmmm what is \C{inMsg.payload[i]} in this case? Is it an object? The example
  demonstrates that we should really have passed in a region that stored other
  transferrable regions inside, i.e. we should proably make the \C{inMsg.payload[i]} a
  region itself. This way we don't have to pay the cost of clone. I think that might be
  an interesting additional discussion in this paragraph. Otherwise I am left as a reader
  with this bittersweat feeling that ``ok you fixed it but you introduced a copy''. But if
  the payload was also on it own transferrable region we would be fine right? We probably
  need a \C{GiveUp} primitive of a region too as we will transfer only the root transferrable
  region (\C{inMsg}) but we want to declare that we've given up ownership on all the nested
  ones too.}

Our experience with Naiad benchmarks suggests that, although our
approach cannot statically enforce memory safety, \dv{Valid point,
  also worth re-visiting in the light of the above?}
it is nonetheless useful, particularly because the static verification comes at no
additional cost to the developer. If the developer is satisfied with
the confidence gained by static verification, she may even choose to
turn the runtime safety checks off. Since we did not implement code
generation in \namec, we could not measure the runtime overhead of
checks needed to enforce safety. However, the number of LOC peforming
operations on \C{Region} objects relative to the total LOC never
exceeds 8\% in Naiad benchmarks. 
% \C{SelectVertex}, for example, contains one each of
% \C{open} and \C{transfer} operations on one region in its 45LOC,
% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
% \C{free} operation.

% We now briefly describe our experience of
% working with these benchmarks. Note that the performance benefits of
% using transferable regions in dataflow systems have been established
% in~\cite{Broom:HotOS}. Our current evaluation only focuses on safety
% aspects.

% \paragraph{Memory safety violation} 


\dv{I belive we are missing a discussion and future work section. This section should
  give sketches of what would it take to integrate other modern language features, e.g. static fields,
  value types? Also another point we could have is discussion on what would it take to integrate this
  work with the garbage collector?}
